The contact value approximation for the pair distribution function of the inhomogeneous hard-sphere fluid

### 2014-05-16 papers/pair-correlation/paper.tex

One common approximation \cite{gloor2007prediction, gross2009density}
assumes a mean-density approximation for the pair distribution
function:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, \tfrac{\pi\sigma^3}{6}\tfrac12
  (n(\rr_1)+n(\rr_2))\right)
\end{align}
where $g(r,\eta)$ is the radial distribution function as a function of
radius and packing fraction $\eta$, and $\sigma$ is the hard-sphere
diameter.  This approach,
however, cannot be applied to highly inhomogeneous systems such as a
dense fluid at a solid surface.  These systems exhibit large density
oscillations that lead to mean packing fractions greater than unity,
which cannot occur in the bulk reference system that defines $g(r,
\eta)$.  The above papers restricted themselves to the liquid-vapor
interface, which does not exhibit this pathology.

Non-pathological approaches use an average of the density over some
volume. Fischer and
Methfessel~\cite{fischer1980born,harris1987comment} introduce the
approximation:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, n_3\left(\tfrac12
  (\rr_1+\rr_2)\right)\right)
  \label{eq:fischer}
\end{align}
where $n_3$ is an integral of the density over a spherical volume that
is now used as one of the fundamental measures in Fundamental
Measure Theory~\cite{rosenfeld1989free}:
\begin{align}
  n_3(\rr) = \int n(\rr')\Theta(\tfrac12 \sigma - |\rr-\rr'|) d\rr'
\end{align}
Equation~\ref{eq:fischer} is computationally awkward, because it
treats as special the midpoint $\tfrac12(\rr_1+\rr_2)$.  Moreover, the
approach of Fischer and Methfessel is intended to approximate the pair
distribution function only at contact, when the distance between
$\rr_1$ and $\rr_2$ is the hard-sphere diameter.

>>>>

One common approximation \cite{gloor2007prediction,
  gross2009density,osborn1980monotonic,mccoy1981comparison,barrett2006some}
assumes a mean-density approximation for the pair distribution
function:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, \tfrac{\pi\sigma^3}{6}\tfrac12
  (n(\rr_1)+n(\rr_2))\right)
\end{align}
where $g(r,\eta)$ is the radial distribution function as a function of
radius and packing fraction $\eta$, and $\sigma$ is the hard-sphere
diameter.  This approach,
however, cannot be applied to highly inhomogeneous systems such as a
dense fluid at a solid surface.  These systems exhibit large density
oscillations that lead to mean packing fractions greater than unity,
which cannot occur in the bulk reference system that defines $g(r,
\eta)$.  The above papers restricted themselves to the liquid-vapor
interface, which does not exhibit this pathology.

Non-pathological approaches use an average of the density over some
volume. Fischer and Methfessel introduce the
approximation~\cite{fischer1980born,harris1987comment,wadewitz2000application,winkelmann2001liquid}:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) \approx g\left(r_{12}, n_3\left(\tfrac12
  (\rr_1+\rr_2)\right)\right)
  \label{eq:fischer}
\end{align}
where $n_3$ is an integral of the density over a spherical volume that
is now used as one of the fundamental measures in Fundamental
Measure Theory~\cite{rosenfeld1989free}:
\begin{align}
  n_3(\rr) = \int n(\rr')\Theta(\tfrac12 \sigma - |\rr-\rr'|) d\rr'
\end{align}
Equation~\ref{eq:fischer} is computationally awkward, because it
treats as special the midpoint $\tfrac12(\rr_1+\rr_2)$.  Moreover, the
approach of Fischer and Methfessel is intended to approximate the pair
distribution function only at contact, when the distance between
$\rr_1$ and $\rr_2$ is the hard-sphere diameter.
